Reanalysis Step 2: Multiple Imputation

### 2025-08-30 backup/reanalysis-2-multiple_imputation_backup_diagnostics.qmd
debug

# Additional packages for imputation
library(mice)
library(miceadds)  # For additional 2-level imputation methods
library(VIM)
library(naniar)

@@

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    cohort_id = factor(cohort_id),  # Ensure cohort_id is a factor
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  ) |>
  mutate(
    # Clean factor levels to remove spaces and special characters for 2l.pmm compatibility
    # Spaces in factor levels cause formula parsing errors in lmer
    # Replace spaces and special characters with underscores
    across(where(is.factor), ~ {
      if (is.factor(.x)) {
        factor(.x, labels = gsub("[^[:alnum:]_]", "_", levels(.x)))
      } else {
        .x
      }
    })
  )
```

@@

# Debug: Check if any methods have invalid values
# Include all standard MICE methods plus 2-level methods from mice and miceadds
valid_methods <- c("", "pmm", "logreg", "polyreg", "polr", "norm", "norm.boot", "norm.nob", 
                   "norm.predict", "sample", "cart", "rf", 
                   # 2-level methods
                   "2l.norm", "2l.lmer", "2l.pan", "2lonly.mean", "2lonly.norm", "2lonly.pmm",
                   # miceadds methods
                   "2l.pmm", "2l.2stage.norm", "2l.2stage.pmm", "ml.lmer")
invalid_methods <- custom_methods[!custom_methods %in% valid_methods]
if (length(invalid_methods) > 0) {
  log_message(paste0("WARNING: Invalid imputation methods detected: ", paste(names(invalid_methods), "=", invalid_methods, collapse=", ")))
  # Reset any invalid methods to appropriate defaults
  for (var in names(invalid_methods)) {
    if (var %in% names(custom_methods)) {
      # Set to empty string to skip imputation for now
      custom_methods[var] <- ""
      log_message(paste0("Reset method for ", var, " to '' (no imputation)"))
    }
  }
}

@@

# Temporal VA variables - use 2-level PMM for clustered data
# 2l.pmm from miceadds: Multilevel PMM using linear mixed models
# This properly accounts for correlation between eyes from same patient
custom_methods["baseline_VA_logMAR"] <- "2l.pmm"  # Multilevel PMM for eyes within patients
custom_methods["intermediate_VA_logMAR"] <- "2l.pmm"  # Multilevel PMM for eyes within patients  
custom_methods["outcome_VA_logMAR"] <- "2l.pmm"  # Multilevel PMM for eyes within patients
custom_methods["cohort_id"] <- ""  # Never impute - data management indicator only

# Other continuous variables - use 2l.pmm for clustered data
custom_methods["CCI"] <- "2l.pmm"  # Multilevel PMM
custom_methods["DCSI"] <- "2l.pmm"  # Multilevel PMM
custom_methods["age"] <- "pmm"  # Regular PMM (patient-level, doesn't vary by eye)

@@

# Mark e_mrn_deidentified as cluster variable
# -2 indicates this is the cluster ID (patient ID) for 2-level imputation
# This is REQUIRED for 2l.pmm to work properly
# The 2l.pmm method from miceadds:
# - Uses linear mixed models (lmer) to model the multilevel structure
# - Preserves within-patient correlation between eyes
# - Obtains PMM donors across all clusters by default
# - Properly accounts for the hierarchical data structure
pred_matrix[, "e_mrn_deidentified"] <- -2
pred_matrix["e_mrn_deidentified", ] <- 0  # Don't predict the cluster ID itself
pred_matrix[, "row_id"] <- 0
pred_matrix["row_id", ] <- 0

@@

# For 2l.pmm variables, use a selective set of predictors to avoid formula length issues
# Identify which variables will use 2l.pmm
vars_2l_pmm <- c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR", "CCI", "DCSI")

# Define core predictors for 2l.pmm models (keep it focused)
core_predictors_2l <- c(
  "ever_lapse_binary",  # Exposure
  "age",                # Key demographic
  "gender_cat",         # Gender
  "baseline_VA_logMAR", # Baseline VA (for temporal)
  "intermediate_VA_logMAR", # Intermediate (for outcome)
  "person_dr",          # DR severity
  "anti_VEGF",          # Key treatment
  "CCI", "DCSI",       # Comorbidities
  "fellow_baseline_VA", # Fellow eye for correlation
  "fellow_outcome_VA"   # Fellow eye outcome
)

# For 2l.pmm variables, limit predictors to core set
log_message("Configuring predictor matrix for 2l.pmm variables with limited predictors")
for (var in vars_2l_pmm) {
  if (var %in% rownames(pred_matrix)) {
    # First set all to 0
    pred_matrix[var, ] <- 0
    # Then selectively enable core predictors
    n_predictors <- 0
    for (pred in core_predictors_2l) {
      if (pred %in% colnames(pred_matrix) && pred != var) {
        pred_matrix[var, pred] <- 1
        n_predictors <- n_predictors + 1
      }
    }
    log_message(paste0("  ", var, ": ", n_predictors, " predictors"))
  }
}

# Ensure exposure is used as predictor for all outcomes
pred_matrix[, "ever_lapse_binary"] <- 1
pred_matrix["ever_lapse_binary", "ever_lapse_binary"] <- 0

@@

>>>>

# Additional packages for imputation
library(mice)
library(VIM)
library(naniar)

@@

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    cohort_id = factor(cohort_id),  # Ensure cohort_id is a factor
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  )
```

@@

# Debug: Check if any methods have invalid values
invalid_methods <- custom_methods[!custom_methods %in% c("", "pmm", "logreg", "polyreg", "polr", "norm", "norm.boot", "norm.nob", "norm.predict", "sample", "cart", "rf", "2l.norm", "2l.lmer", "2l.pan", "2lonly.mean", "2lonly.norm", "2lonly.pmm")]
if (length(invalid_methods) > 0) {
  log_message(paste0("WARNING: Invalid imputation methods detected: ", paste(names(invalid_methods), "=", invalid_methods, collapse=", ")))
  # Reset any invalid methods to appropriate defaults
  for (var in names(invalid_methods)) {
    if (var %in% names(custom_methods)) {
      # Set to empty string to skip imputation for now
      custom_methods[var] <- ""
      log_message(paste0("Reset method for ", var, " to '' (no imputation)"))
    }
  }
}

@@

# Temporal VA variables - use 2-level PMM to account for patient clustering
custom_methods["baseline_VA_logMAR"] <- "2l.pmm"  # 2-level PMM for clustered data
custom_methods["intermediate_VA_logMAR"] <- "2l.pmm"  # 2-level PMM for clustered data
custom_methods["outcome_VA_logMAR"] <- "2l.pmm"  # 2-level PMM for clustered data
custom_methods["cohort_id"] <- ""  # Never impute - data management indicator only

# Other continuous variables - use 2-level PMM for patient clustering
custom_methods["CCI"] <- "2l.pmm"  # 2-level PMM for count data
custom_methods["DCSI"] <- "2l.pmm"  # 2-level PMM for count data
custom_methods["age"] <- "2l.pmm"  # Keep as regular PMM since age doesn't vary within patient

@@

# Mark e_mrn_deidentified as cluster variable for 2-level imputation
# -2 indicates this is the cluster ID (patient ID)
pred_matrix[, "e_mrn_deidentified"] <- -2
pred_matrix["e_mrn_deidentified", ] <- 0  # Don't predict the cluster ID itself
pred_matrix[, "row_id"] <- 0
pred_matrix["row_id", ] <- 0

@@

# Ensure exposure is used as predictor for outcomes
pred_matrix[, "ever_lapse_binary"] <- 1
pred_matrix["ever_lapse_binary", "ever_lapse_binary"] <- 0

@@

## Extract Complete Datasets

**Note:** In the stripplots below, the leftmost column (labeled 1) shows the **original observed data** with missing values. Columns 2 through `r n_imputations + 1` show the imputed datasets.

```{r imputation-diagnostics}
# Check imputed values for key variables
# In mice stripplots: .imp = 1 is original data, .imp = 2+ are imputations

# CCI - compare observed vs imputed
if(sum(is.na(df_mice$CCI)) > 0) {
  # Density plot: Blue = observed data, Red = imputed values
  densityplot(mice_results, ~ CCI, 
              main = "Charlson Comorbidity Index (CCI)") |> print()

  # Stripplot: Column 1 = original data, Columns 2+ = imputed datasets
  stripplot(mice_results, CCI ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "CCI Distribution: Original vs Imputed") |> print()
}

# DCSI - compare observed vs imputed  
if(sum(is.na(df_mice$DCSI)) > 0) {
  densityplot(mice_results, ~ DCSI,
              main = "Diabetes Complications Severity Index (DCSI)") |> print()

  stripplot(mice_results, DCSI ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "DCSI Distribution: Original vs Imputed") |> print()
}

# Baseline VA - compare observed vs imputed
if(sum(is.na(df_mice$baseline_VA_logMAR)) > 0) {
  densityplot(mice_results, ~ baseline_VA_logMAR,
              main = "Baseline Visual Acuity (logMAR)") |> print()

  stripplot(mice_results, baseline_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "Baseline VA Distribution: Original vs Imputed") |> print()
}

# Intermediate VA - compare observed vs imputed
if(sum(is.na(df_mice$intermediate_VA_logMAR)) > 0) {
  densityplot(mice_results, ~ intermediate_VA_logMAR,
              main = "Intermediate Visual Acuity (logMAR)") |> print()

  stripplot(mice_results, intermediate_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "Intermediate VA Distribution: Original vs Imputed") |> print()
}

# Outcome VA - compare observed vs imputed
if(sum(is.na(df_mice$outcome_VA_logMAR)) > 0) {
  densityplot(mice_results, ~ outcome_VA_logMAR,
              main = "Outcome Visual Acuity at 2 Years (logMAR)") |> print()

  stripplot(mice_results, outcome_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "2-Year Outcome VA Distribution: Original vs Imputed") |> print()
}

# Person DR - compare observed vs imputed (categorical)
if(sum(is.na(df_mice$person_dr)) > 0) {
  # For categorical variables, check proportions across imputations
  # Extract each imputation and tabulate
  dr_props <- sapply(1:n_imputations, function(i) {
    imp_data <- complete(mice_results, i)
    prop.table(table(imp_data$person_dr))
  })

  # Create summary table
  dr_summary <- data.frame(
    Category = c("No_DR", "NPDR", "PDR"),
    Mean_Proportion = c(round(mean(dr_props[1,]), 3), 
                       round(mean(dr_props[2,]), 3),
                       round(mean(dr_props[3,]), 3)),
    SD = c(round(sd(dr_props[1,]), 3),
           round(sd(dr_props[2,]), 3), 
           round(sd(dr_props[3,]), 3))
  )
  kable(dr_summary, caption = "DR severity proportions across imputations")

  # Visual check if stripplot works with factors
  tryCatch({
    stripplot(mice_results, person_dr ~ .imp, pch = 20, cex = 1.2) |> print()
  }, error = function(e) {
    # Note: Stripplot not available for categorical person_dr
  })
}

# Check imputation quality for categorical variables
# Cataract surgery
if(sum(is.na(df_mice$catsurg_before_hitplus_cat)) > 0) {
  # Extract proportions for each imputation
  cat_props <- sapply(1:n_imputations, function(i) {
    imp_data <- complete(mice_results, i)
    prop.table(table(imp_data$catsurg_before_hitplus_cat))
  })

  # Create summary table for cataract surgery proportions
  if(is.matrix(cat_props)) {
    cat_summary <- data.frame(
      Level = rownames(cat_props),
      Mean_Proportion = apply(cat_props, 1, mean) |> round(3),
      SD = apply(cat_props, 1, sd) |> round(3)
    )
  } else {
    cat_summary <- data.frame(
      Level = "Cataract Surgery",
      Mean_Proportion = round(mean(cat_props), 3),
      SD = round(sd(cat_props), 3)
    )
  }
  kable(cat_summary, caption = "Cataract surgery proportions across imputations")
}

# Fellow Eye Correlation Check
# For patients with both eyes, check correlation between eyes
if (sum(df_mice$n_eyes == 2, na.rm = TRUE) > 0) {

  # Extract first imputed dataset
  imp_data_1 <- complete(mice_results, 1)

  # Get bilateral patients
  bilateral_patients <- imp_data_1 |>
    filter(n_eyes == 2) |>
    arrange(e_mrn_deidentified)  # Sort by patient ID

  # Check correlation for baseline VA
  # Use slice_head and slice_tail to get first and second eye for each patient
  baseline_pairs <- bilateral_patients |>
    group_by(e_mrn_deidentified) |>
    summarise(
      eye1_baseline = first(baseline_VA_logMAR),
      eye2_baseline = last(baseline_VA_logMAR),
      .groups = "drop"
    ) |>
    filter(!is.na(eye1_baseline) & !is.na(eye2_baseline))

  baseline_cor <- cor(baseline_pairs$eye1_baseline, baseline_pairs$eye2_baseline)

  # Check correlation for outcome VA 
  outcome_pairs <- bilateral_patients |>
    group_by(e_mrn_deidentified) |>
    summarise(
      eye1_outcome = first(outcome_VA_logMAR),
      eye2_outcome = last(outcome_VA_logMAR),
      .groups = "drop"
    ) |>
    filter(!is.na(eye1_outcome) & !is.na(eye2_outcome))

  outcome_cor <- cor(outcome_pairs$eye1_outcome, outcome_pairs$eye2_outcome)

  # Summary table
  fellow_cor_summary <- data.frame(
    Measure = c("Baseline VA", "Outcome VA"),
    N_Pairs = c(nrow(baseline_pairs), nrow(outcome_pairs)),
    Correlation = c(round(baseline_cor, 3), round(outcome_cor, 3)),
    Expected_Range = c("0.7-0.9", "0.7-0.9"),
    Status = c(
      ifelse(baseline_cor >= 0.7 & baseline_cor <= 0.9, "✓ Within expected", "⚠ Outside expected"),
      ifelse(outcome_cor >= 0.7 & outcome_cor <= 0.9, "✓ Within expected", "⚠ Outside expected")
    )
  )

  kable(fellow_cor_summary, caption = "Fellow eye correlation in imputed data")

  # Visual check - scatter plot
  par(mfrow = c(1, 2))

  # Baseline VA correlation plot
  plot(baseline_pairs$eye1_baseline, baseline_pairs$eye2_baseline,
       xlab = "Eye 1 Baseline VA (logMAR)",
       ylab = "Eye 2 Baseline VA (logMAR)",
       main = paste0("Baseline VA Correlation (r = ", round(baseline_cor, 2), ")"),
       pch = 19, col = rgb(0, 0, 0, 0.5))
  abline(lm(eye2_baseline ~ eye1_baseline, data = baseline_pairs), col = "red", lwd = 2)

  # Outcome VA correlation plot
  plot(outcome_pairs$eye1_outcome, outcome_pairs$eye2_outcome,
       xlab = "Eye 1 Outcome VA (logMAR)",
       ylab = "Eye 2 Outcome VA (logMAR)",
       main = paste0("Outcome VA Correlation (r = ", round(outcome_cor, 2), ")"),
       pch = 19, col = rgb(0, 0, 0, 0.5))
  abline(lm(eye2_outcome ~ eye1_outcome, data = outcome_pairs), col = "red", lwd = 2)

  par(mfrow = c(1, 1))

  # Check correlation across all imputations
  all_cors <- sapply(1:n_imputations, function(i) {
    imp_data <- complete(mice_results, i)
    bilateral <- imp_data |> filter(n_eyes == 2)

    # Baseline correlation
    base_pairs <- bilateral |>
      group_by(e_mrn_deidentified) |>
      summarise(
        eye1 = first(baseline_VA_logMAR),
        eye2 = last(baseline_VA_logMAR),
        .groups = "drop"
      ) |>
      filter(!is.na(eye1) & !is.na(eye2))

    base_cor <- if(nrow(base_pairs) > 0) cor(base_pairs$eye1, base_pairs$eye2) else NA

    # Outcome correlation
    out_pairs <- bilateral |>
      group_by(e_mrn_deidentified) |>
      summarise(
        eye1 = first(outcome_VA_logMAR),
        eye2 = last(outcome_VA_logMAR),
        .groups = "drop"
      ) |>
      filter(!is.na(eye1) & !is.na(eye2))

    out_cor <- if(nrow(out_pairs) > 0) cor(out_pairs$eye1, out_pairs$eye2) else NA

    c(baseline_cor = base_cor, outcome_cor = out_cor)
  })

  # Summary of correlations across imputations
  cor_summary_all <- data.frame(
    Measure = c("Baseline VA", "Outcome VA"),
    Mean_Correlation = c(
      round(mean(all_cors["baseline_cor",], na.rm = TRUE), 3),
      round(mean(all_cors["outcome_cor",], na.rm = TRUE), 3)
    ),
    SD_Correlation = c(
      round(sd(all_cors["baseline_cor",], na.rm = TRUE), 3),
      round(sd(all_cors["outcome_cor",], na.rm = TRUE), 3)
    ),
    Min = c(
      round(min(all_cors["baseline_cor",], na.rm = TRUE), 3),
      round(min(all_cors["outcome_cor",], na.rm = TRUE), 3)
    ),
    Max = c(
      round(max(all_cors["baseline_cor",], na.rm = TRUE), 3),
      round(max(all_cors["outcome_cor",], na.rm = TRUE), 3)
    )
  )

  kable(cor_summary_all, caption = "Fellow eye correlations across all imputations")
}

# Clinical Plausibility Check for Imputed Values
log_message("Checking clinical plausibility of imputed values")

# Extract observed and imputed ranges for each variable
plausibility_checks <- list()

# Visual Acuity measures (logMAR scale)
# Clinical limits: -0.3 (exceptional vision) to 3.0 (hand motion/counting fingers)
va_vars <- c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR")

for (var in va_vars) {
  if (sum(is.na(df_mice[[var]])) > 0) {
    # Get observed range
    obs_range <- range(df_mice[[var]], na.rm = TRUE)

    # Get imputed ranges across all imputations
    imp_ranges <- sapply(1:n_imputations, function(i) {
      imp_data <- complete(mice_results, i)
      # Only look at previously missing values
      was_missing <- is.na(df_mice[[var]])
      if (sum(was_missing) > 0) {
        range(imp_data[[var]][was_missing])
      } else {
        c(NA, NA)
      }
    })

    plausibility_checks[[var]] <- data.frame(
      Variable = var,
      Observed_Min = round(obs_range[1], 3),
      Observed_Max = round(obs_range[2], 3),
      Imputed_Min = round(min(imp_ranges[1,], na.rm = TRUE), 3),
      Imputed_Max = round(max(imp_ranges[2,], na.rm = TRUE), 3),
      Clinical_Min = -0.3,
      Clinical_Max = 3.0,
      stringsAsFactors = FALSE
    )

    # Check if imputed values fall within observed range
    plausibility_checks[[var]]$Within_Observed <- 
      plausibility_checks[[var]]$Imputed_Min >= plausibility_checks[[var]]$Observed_Min &
      plausibility_checks[[var]]$Imputed_Max <= plausibility_checks[[var]]$Observed_Max

    # Check if imputed values fall within clinical limits
    plausibility_checks[[var]]$Within_Clinical <- 
      plausibility_checks[[var]]$Imputed_Min >= plausibility_checks[[var]]$Clinical_Min &
      plausibility_checks[[var]]$Imputed_Max <= plausibility_checks[[var]]$Clinical_Max
  }
}

# Comorbidity indices (non-negative integers)
comorbidity_vars <- c("CCI", "DCSI")

for (var in comorbidity_vars) {
  if (sum(is.na(df_mice[[var]])) > 0) {
    obs_range <- range(df_mice[[var]], na.rm = TRUE)

    imp_ranges <- sapply(1:n_imputations, function(i) {
      imp_data <- complete(mice_results, i)
      was_missing <- is.na(df_mice[[var]])
      if (sum(was_missing) > 0) {
        range(imp_data[[var]][was_missing])
      } else {
        c(NA, NA)
      }
    })

    plausibility_checks[[var]] <- data.frame(
      Variable = var,
      Observed_Min = round(obs_range[1], 1),
      Observed_Max = round(obs_range[2], 1),
      Imputed_Min = round(min(imp_ranges[1,], na.rm = TRUE), 1),
      Imputed_Max = round(max(imp_ranges[2,], na.rm = TRUE), 1),
      Clinical_Min = 0,
      Clinical_Max = ifelse(var == "CCI", 37, 13),  # CCI max 37, DCSI max 13
      stringsAsFactors = FALSE
    )

    plausibility_checks[[var]]$Within_Observed <- 
      plausibility_checks[[var]]$Imputed_Min >= plausibility_checks[[var]]$Observed_Min &
      plausibility_checks[[var]]$Imputed_Max <= plausibility_checks[[var]]$Observed_Max

    plausibility_checks[[var]]$Within_Clinical <- 
      plausibility_checks[[var]]$Imputed_Min >= plausibility_checks[[var]]$Clinical_Min &
      plausibility_checks[[var]]$Imputed_Max <= plausibility_checks[[var]]$Clinical_Max
  }
}

# Combine all plausibility checks
if (length(plausibility_checks) > 0) {
  plausibility_summary <- do.call(rbind, plausibility_checks)
  rownames(plausibility_summary) <- NULL

  # Add status column
  plausibility_summary$Status <- ifelse(
    plausibility_summary$Within_Observed,
    "✓ Within observed range (PMM working correctly)",
    ifelse(
      plausibility_summary$Within_Clinical,
      "⚠ Outside observed but within clinical limits",
      "✗ Outside clinical limits"
    )
  )

  kable(plausibility_summary[, c("Variable", "Observed_Min", "Observed_Max", 
                                 "Imputed_Min", "Imputed_Max", "Status")],
        caption = "Clinical Plausibility of Imputed Values (PMM ensures values stay within observed range)")

  log_message(paste0("Plausibility check complete. All variables within observed range: ",
                    all(plausibility_summary$Within_Observed)))
}

# Check between-imputation variance
log_message("Checking between-imputation variance")

# Calculate variance across imputations for key outcomes
variance_checks <- sapply(va_vars, function(var) {
  if (sum(is.na(df_mice[[var]])) > 0) {
    # Get imputed values for each missing observation across all imputations
    was_missing <- which(is.na(df_mice[[var]]))

    if (length(was_missing) > 0) {
      imp_values <- sapply(1:n_imputations, function(i) {
        imp_data <- complete(mice_results, i)
        imp_data[[var]][was_missing]
      })

      # Calculate mean variance across observations
      if (is.matrix(imp_values)) {
        obs_variances <- apply(imp_values, 1, var)
        mean_var <- mean(obs_variances, na.rm = TRUE)
        min_var <- min(obs_variances, na.rm = TRUE)
        max_var <- max(obs_variances, na.rm = TRUE)
      } else {
        mean_var <- var(imp_values)
        min_var <- mean_var
        max_var <- mean_var
      }

      c(Mean_Variance = mean_var, Min_Variance = min_var, Max_Variance = max_var)
    } else {
      c(Mean_Variance = NA, Min_Variance = NA, Max_Variance = NA)
    }
  } else {
    c(Mean_Variance = NA, Min_Variance = NA, Max_Variance = NA)
  }
})

if (ncol(variance_checks) > 0) {
  variance_summary <- data.frame(
    Variable = colnames(variance_checks),
    Mean_Between_Imp_Variance = round(variance_checks["Mean_Variance",], 4),
    Min_Between_Imp_Variance = round(variance_checks["Min_Variance",], 4),
    Max_Between_Imp_Variance = round(variance_checks["Max_Variance",], 4),
    Status = ifelse(
      variance_checks["Mean_Variance",] > 0.001,
      "✓ Adequate variance",
      "⚠ Low variance - check predictor weights"
    )
  )

  kable(variance_summary, 
        caption = "Between-Imputation Variance (ensures uncertainty is preserved)")

  log_message(paste0("Variance check complete. Adequate variance in all variables: ",
                    all(variance_checks["Mean_Variance",] > 0.001, na.rm = TRUE)))
}
```

## MNAR Sensitivity Analysis: Delta-Shift Method

```{r mnar-sensitivity}
# Delta-shift sensitivity analysis for MNAR
# This shows how results would change if missing values were systematically different
# We do NOT apply these shifts to the main imputation - just show diagnostics

log_message("Performing MNAR sensitivity analysis (diagnostic only)")

# Define small symmetric deltas for outcome VA (in logMAR units)
# 0.1 logMAR = 1 line on eye chart
deltas <- c(-0.2, -0.1, 0, 0.1, 0.2)

# For each delta, show how the outcome distribution would shift
delta_results <- list()

for (delta in deltas) {
  # Take first imputed dataset as example
  imp_data_1 <- complete(mice_results, 1)

  # Identify which outcomes were imputed
  was_imputed <- is.na(df_mice$outcome_VA_logMAR)

  if (sum(was_imputed) > 0) {
    # Create shifted version (for diagnostic only)
    shifted_data <- imp_data_1
    shifted_data$outcome_VA_logMAR[was_imputed] <- 
      shifted_data$outcome_VA_logMAR[was_imputed] + delta

    # Calculate outcome statistics
    delta_results[[as.character(delta)]] <- data.frame(
      Delta = delta,
      Delta_Lines = round(delta / 0.1, 1),  # Convert to eye chart lines
      Mean_Observed = round(mean(imp_data_1$outcome_VA_logMAR[!was_imputed], na.rm = TRUE), 3),
      Mean_Imputed_Original = round(mean(imp_data_1$outcome_VA_logMAR[was_imputed], na.rm = TRUE), 3),
      Mean_Imputed_Shifted = round(mean(shifted_data$outcome_VA_logMAR[was_imputed], na.rm = TRUE), 3),
      Pct_VI_Observed = round(100 * mean(imp_data_1$outcome_VA_logMAR[!was_imputed] >= 1.0, na.rm = TRUE), 1),
      Pct_VI_Imputed_Original = round(100 * mean(imp_data_1$outcome_VA_logMAR[was_imputed] >= 1.0, na.rm = TRUE), 1),
      Pct_VI_Imputed_Shifted = round(100 * mean(shifted_data$outcome_VA_logMAR[was_imputed] >= 1.0, na.rm = TRUE), 1)
    )
  }
}

if (length(delta_results) > 0) {
  sensitivity_summary <- do.call(rbind, delta_results)
  rownames(sensitivity_summary) <- NULL

  # Add interpretation
  sensitivity_summary$Interpretation <- c(
    "Imputed have 2 lines better vision",
    "Imputed have 1 line better vision",
    "No systematic difference (MAR)",
    "Imputed have 1 line worse vision",
    "Imputed have 2 lines worse vision"
  )

  kable(sensitivity_summary[, c("Delta", "Delta_Lines", "Interpretation",
                                "Mean_Imputed_Original", "Mean_Imputed_Shifted",
                                "Pct_VI_Imputed_Original", "Pct_VI_Imputed_Shifted")],
        caption = "MNAR Sensitivity Analysis: Effect of systematic differences in missing outcomes",
        col.names = c("Delta (logMAR)", "Lines on chart", "Interpretation",
                     "Mean VA (Original)", "Mean VA (Shifted)",
                     "% VI (Original)", "% VI (Shifted)"))

  # Create visualization
  par(mfrow = c(1, 2))

  # Plot 1: Mean VA by delta
  plot(sensitivity_summary$Delta, sensitivity_summary$Mean_Imputed_Shifted,
       type = "b", pch = 19,
       xlab = "Delta adjustment (logMAR)",
       ylab = "Mean imputed outcome VA (logMAR)",
       main = "Sensitivity to MNAR assumption",
       ylim = range(c(sensitivity_summary$Mean_Observed,
                     sensitivity_summary$Mean_Imputed_Shifted)))
  abline(h = sensitivity_summary$Mean_Observed[1], col = "blue", lty = 2)
  abline(v = 0, col = "gray", lty = 3)
  legend("topleft", c("Shifted imputed", "Observed mean"), 
         col = c("black", "blue"), lty = c(1, 2), pch = c(19, NA))

  # Plot 2: % VI by delta
  plot(sensitivity_summary$Delta, sensitivity_summary$Pct_VI_Imputed_Shifted,
       type = "b", pch = 19,
       xlab = "Delta adjustment (logMAR)",
       ylab = "% with vision impairment",
       main = "VI prevalence under MNAR",
       ylim = range(c(sensitivity_summary$Pct_VI_Observed,
                     sensitivity_summary$Pct_VI_Imputed_Shifted)))
  abline(h = sensitivity_summary$Pct_VI_Observed[1], col = "blue", lty = 2)
  abline(v = 0, col = "gray", lty = 3)
  legend("topleft", c("Shifted imputed", "Observed prevalence"), 
         col = c("black", "blue"), lty = c(1, 2), pch = c(19, NA))

  par(mfrow = c(1, 1))

  log_message("MNAR sensitivity analysis complete (diagnostic only - main imputation unchanged)")
}
```

### 2025-08-30 reanalysis-2-multiple_imputation.qmd
debug

---
title: "Reanalysis Step 2: Multiple Imputation (Clean Version)"
subtitle: "Simplified imputation using mice with clustering"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

# Initialize logging
log_file <- init_log("multiple_imputation_clean")

@@

log_message("Imputation methods configured")
```

@@

# Save convergence plot
pdf(file.path(reanalysis_data_dir, "mice_convergence_clean.pdf"))
plot(imp)
dev.off()
```

@@

# Save the mice object
saveRDS(imp, file.path(reanalysis_data_dir, "mice_results_clean.rds"))
log_message(paste0("Saved MICE object to: mice_results_clean.rds"))

# Save the list of complete datasets
saveRDS(imputed_list, file.path(reanalysis_data_dir, "imputed_datasets_clean.rds"))
log_message(paste0("Saved ", length(imputed_list), " imputed datasets"))

>>>>

---
title: "Reanalysis Step 2: Multiple Imputation"
subtitle: "Imputation using mice with strict temporal constraints"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

# Initialize logging
log_file <- init_log("multiple_imputation")

@@

log_message("Imputation methods configured")

# Print methods summary to verify configuration
methods_summary <- data.frame(
  Variable = names(methods),
  Method = methods,
  stringsAsFactors = FALSE
) |>
  mutate(
    Will_Impute = Method != ""
  )

log_message("\nVariables that WILL be imputed:")
print(methods_summary |> filter(Will_Impute))

log_message("\nVariables that will NOT be imputed:")
print(methods_summary |> filter(!Will_Impute))

# Specifically check our key variables
key_vars_check <- c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR", "person_dr")
log_message("\nMethods for key variables with missingness:")
for (var in key_vars_check) {
  if (var %in% names(methods)) {
    log_message(paste0(var, ": ", methods[var]))
  } else {
    log_message(paste0(var, ": NOT IN DATASET"))
  }
}
```

@@

# Verify imputation results
log_message("\nVerifying imputation results...")
first_imp <- complete(imp, 1)
log_message("Missing data in first imputed dataset:")
missing_after <- colSums(is.na(first_imp))
missing_after <- missing_after[missing_after > 0]
if (length(missing_after) > 0) {
  print(missing_after)
  log_message("WARNING: The above variables still have missing data after imputation!")
} else {
  log_message("SUCCESS: All missing data has been imputed.")
}

# Save convergence plot
pdf(file.path(reanalysis_data_dir, "mice_convergence.pdf"))
plot(imp)
dev.off()
```

@@

# Save the mice object
saveRDS(imp, file.path(reanalysis_data_dir, "mice_results.rds"))
log_message(paste0("Saved MICE object to: mice_results.rds"))

# Save the list of complete datasets
saveRDS(imputed_list, file.path(reanalysis_data_dir, "imputed_datasets.rds"))
log_message(paste0("Saved ", length(imputed_list), " imputed datasets"))
